STA302 Fall 2024 Final Project Part 1: Research Proposal and Data Introduction

### 2024-11-25 final_paper/paper.qmd
Added final report

# Preliminary Results 
Our model examines the relationship between the New Housing Price Index (NHPI) and predictors like detached absorption, unabsorbed homes, new homes under construction, completed constructions, and rate hikes. The intercept is 88.52, representing the NHPI value when all predictors are zero, serving as a baseline for evaluating the effects of other variables.

Detached absorption, our first predictor, shows that each additional home sold in a quarter is associated with a decrease of 0.0227 points in the NHPI. Conversely, unabsorbed homes increase the NHPI by 0.005315 points. The impact of new homes under construction is minimal, increasing the NHPI by 0.0009611 points for each home started. Completed constructions have a greater effect, contributing a 0.002977 point increase. Rate hikes, however, have the most significant influence: each rate increase leads to a 13.41-point decrease in the NHPI.

## Residual Analysis
We assessed the assumptions of our linear regression model by examining residual plots. The Residuals vs. Fitted Values plot (@fig-resid) reveals clustering around a fitted value of 100, with a slight decrease in spread, indicating potential non-linearity and correlated errors.

Further, we reviewed residuals vs. predictors. The residual plots for homes under construction and completed homes (@fig-residuals-3 and @fig-residuals-4) show minimal violations, as the spread appears random and well-distributed. However, the residuals for sold (@fig-residuals-1) and unsold homes (@fig-residuals-2) display fanning and clustering, suggesting correlated errors and non-constant variance. Additionally, the categorical predictor "rate hike" (@fig-residuals-5) shows wider residual variance for a value of "1" than for "0," indicating non-constant variance.

Lastly, while the QQ plot (@fig-qq_plot) suggests that most residuals align with the expected normal distribution, deviations at the lower and upper extremes indicate some skewness. This suggests that the residuals are not perfectly normally distributed, hinting at a potential violation of the normality assumption.

## Model Fitting {#sec-model}

**Data**  
- **Quarterly Average**: This is our response variable, representing the average cost of a home in Canada.  
- **Detached Absorption Quarterly Avg**: This is our first predictor variable, describing the number of houses sold in a quarter.  
- **Starting Detached Construction**:  This predictor variable indicates the number of houses that began construction in a quarter.  
- **Completed Construction Detached**:  This represents the number of homes completed in a given quarter.  
- **Rates**:  This is our categorical predictor, with a value of 1 if there was a rate hike in that quarter or 0 if there was not.  
- **Detached Unabsorbed Quarterly Avg**:  This is our final predictor, indicating the number of available houses that were not sold during that quarter.

```{r}
#| label: fig-data
#| fig-cap: "Data"
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(readr)

# Load the dataset into R
data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```

```{r}
#| label: fig-model
#| fig-cap: "Multiple Linear Model"
#| echo: false
#| warning: false
#| message: false

library(tibble)
library(knitr)

# multiple linear model
model <- lm(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg + 
              Detached_Unabsorbed_Quarterly_Avg + 
              Starting_Detached_Construction + 
              Completed_Construction_Detached + 
              rates, 
            data = data)

# create a data frame
coefficients_table <- as.data.frame(coef(model))
coefficients_table <- coefficients_table %>%
  rownames_to_column("Variable") %>%
  rename("Coefficient" = "coef(model)")

coefficients_table <- coefficients_table[1:6, ]

# summary table
kable(coefficients_table, caption = "Summary Table of Regression Coefficients", format = "markdown")
```

```{r}
#| label: fig-resid
#| fig-cap: "Residual vs Fitted Values"
#| echo: false
#| warning: false
#| message: false

# Residual and fitted values for the model
y_value <- resid(model)
x_value <- fitted(model)

# Plot the residual vs fitted values
plot(x = x_value, y = y_value, 
     main = "Residual vs Fitted", 
     xlab = "Fitted", 
     ylab = "Residuals")
```

```{r}
#| label: fig-residuals
#| fig-cap: "Residuals vs Each Predictor"
#| echo: false
#| warning: false
#| message: false

# Set up plot layout: 1 row, 3 columns
par(mfrow = c(1, 1))

# Residual vs Sold Homes
plot(x = data$Detached_Absorption_Quarterly_Avg, y = y_value, 
     main = "Residual vs Sold Homes", 
     xlab = "Sold Homes", ylab = "Residuals")

# Residual vs Unsold Homes
plot(x = data$Detached_Unabsorbed_Quarterly_Avg, y = y_value, 
     main = "Residual vs Unsold Homes", 
     xlab = "Unsold Homes", ylab = "Residuals")

# Residual vs New Homes Being Built
plot(x = data$Starting_Detached_Construction, y = y_value, 
     main = "Residual vs New Homes Being Built", 
     xlab = "New Homes Being Built", ylab = "Residuals")

# Residual vs Homes Completed
plot(x = data$Completed_Construction_Detached, y = y_value, 
     main = "Residual vs Homes Completed", 
     xlab = "Homes Completed", ylab = "Residuals")

# Boxplot of Residuals vs Rate Hike
boxplot(y_value ~ data$rates, 
        main = "Residual vs Rate Hike", 
        xlab = "Rate Hike", ylab = "Residuals")
```

```{r}
#| label: fig-qq_plot
#| fig-cap: "QQ Plot"
#| echo: false
#| warning: false
#| message: false

# Create a QQ plot for residuals
qqnorm(resid(model), main = "Normal QQ Plot of Residuals")
qqline(resid(model))
```

>>>>

# Preliminary Results

Problem Definition
Focusing on the *New Housing Price Index* (NHPI) as the response variable, We will analyze economic, construction, and financial predictors such as construction activity, interest rates, and GDP.

## **1. Data Preparation**

- **Tools Used**: Analysis was conducted in **R** using `tidyverse` for data manipulation and `car` for diagnostics.

- **Steps Taken**:  
  - Defined **New Housing Price Index (NHPI)** as the response variable and identified predictors:  
  1. **Detached Absorption (Quarterly Average)**  
  2. **Semi Absorption (Quarterly Average)**  
  3. **Detached Unabsorbed Homes (Quarterly Average)**  
  4. **Semi Unabsorbed Homes (Quarterly Average)**  
  5. **GDP (Quarterly Average)**  
  6. **CPI (Quarterly Average)**  
  7. **Starting Detached Construction**  
  8. **Starting Semi-Detached Construction**  
  9. **Under Construction Detached Homes**  
  10. **Under Construction Semi-Detached Homes**  
  11. **Completed Construction Semi-Detached**  
  12. **Completed Construction Detached**  
  13. **Rate Hikes**
  - Removed missing values and outliers using Cook’s Distance.
  - Merged datasets from 1997–2016, ensuring quarterly granularity.  
  - **Split the Data**:  
    - Divided the dataset into training and test/validation sets.  
    - Used a common split, such as 50/50, to ensure both subsets are sufficiently large.  
    - Performed the split randomly to avoid sampling bias.  
```{r}
# Load necessary libraries
library(tidyverse)
library(here)
library(caret)

# Load the dataset
data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

# response variable and predictors
response <- "Quarterly_Average"
predictors <- c(
  "Detached_Absorption_Quarterly_Avg",
  "Semi_Absorption_Quarterly_Avg",
  "Detached_Unabsorbed_Quarterly_Avg",
  "Semi_Unabsorbed_Quarterly_Avg",
  "GDP_Quarterly_Avg",
  "CPI_Quarterly_Avg",
  "Starting_Detached_Construction",
  "Starting_Semi_Construction",
  "Under_Construction_Detached",
  "Under_Construction_Semi",
  "Completed_Construction_Semi",
  "Completed_Construction_Detached",
  "rates"
)

# Subset the relevant data
selected_data <- data %>%
  dplyr::select(c(response, all_of(predictors)))

# Clean Data

## Handle missing values
cleaned_data <- selected_data %>%
  drop_na()

## Remove duplicates
cleaned_data <- cleaned_data %>%
  distinct()

# Fit an initial regression model
model <- lm(as.formula(paste(response, "~", paste(predictors, collapse = " + "))), data = cleaned_data)

# Calculate Cook's Distance
cooks_d <- cooks.distance(model)

# Identify influential points
influential_points <- which(cooks_d > 4 / (nrow(cleaned_data) - length(model$coefficients)))

# Exclude influential observations
cleaned_data <- cleaned_data[-influential_points, ]

# Split the Data
set.seed(123) # Ensure reproducibility
train_index <- createDataPartition(cleaned_data[[response]], p = 0.5, list = FALSE)
train_data <- cleaned_data[train_index, ]
test_data <- cleaned_data[-train_index, ]

list(
  training_set = train_data,
  test_set = test_data
)

```

## **2. Exploratory Data Analysis (EDA)**
- **Tools Used**: `ggplot2` for visualization and `corrplot` for examining correlations.  
- **Key Insights**:  
  - **Scatterplots** indicated potential linear relationships between predictors and NHPI.  
  - **Variance Inflation Factors (VIFs)** flagged multicollinearity among construction-related predictors, necessitating feature selection or transformation.  
  - Outliers were identified, influencing NHPI trends, and removed to enhance model accuracy.  
  - **Explore Training and Test Sets**:  
    - Performed EDA on both `training_set` and `test_set` to ensure consistency and reliability.  
    - Compared distributions, means, variances, and shapes of variables across both sets to ensure similarity.  
    - Checked for any skews, extreme observations, or discrepancies between the training and test sets that could impact model generalizability.

### Train Data
```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(car)       # For assumption checks like residual plots
library(corrplot)  # For correlation matrix visualization
library(broom)     # For augmenting model output
library(dplyr)

# Define predictors and response variable
response <- "Quarterly_Average"
predictors <- setdiff(names(train_data), c("Quarter", response))

# Subset relevant data from the training set
eda_training_data <- train_data %>%
  select(all_of(c(response, predictors)))

### **1. Visualize Relationships in Training Set** ###
# Scatterplots for predictors in the training set
for (predictor in predictors) {
  p <- ggplot(eda_training_data, aes_string(x = predictor, y = response)) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", color = "blue", se = FALSE) +
    labs(title = paste("Scatterplot of", response, "vs", predictor, "(Training Set)"),
         x = predictor, y = response) +
    theme_minimal()

  print(p)  # Explicitly print the plot
}

### **2. Correlation Analysis in Training Set** ###
# Correlation matrix for the training set
cor_matrix_training <- cor(eda_training_data, use = "complete.obs")
print(cor_matrix_training)
corrplot(cor_matrix_training, method = "circle", type = "upper", tl.col = "black", tl.srt = 45)

### **3. Fit Linear Regression Model on Training Set** ###
# Fit the model using the training set
model_training <- lm(as.formula(paste(response, "~", paste(predictors, collapse = " + "))), data = eda_training_data)
summary(model_training)

# Extract residuals and fitted values from the training model
residuals_training <- resid(model_training)
fitted_values_training <- fitted(model_training)

# Calculate Variance Inflation Factors (VIF) for the training model
vif_training <- vif(model_training)
print(vif_training)

### **4. Outlier Detection in Training Set** ###
# Identify influential points using Cook's Distance
cooks_d_training <- cooks.distance(model_training)
influential_points_training <- which(cooks_d_training > 4 / (nrow(eda_training_data) - length(model_training$coefficients)))

# Output details of influential points
if (length(influential_points_training) > 0) {
  print(paste("Number of influential points:", length(influential_points_training)))
  print("Influential points identified:")
  print(influential_points_training)
} else {
  print("No influential points detected.")
}

```
### Test Data
```{r}
# Load necessary libraries
library(tidyverse)
library(ggplot2)
library(car)       # For assumption checks like residual plots
library(corrplot)  # For correlation matrix visualization
library(broom)     # For augmenting model output
library(dplyr)

# Define predictors and response variable
response <- "Quarterly_Average"
predictors <- setdiff(names(test_data), c("Quarter", response))

# Subset relevant data from the test set
eda_test_data <- test_data %>%
  select(all_of(c(response, predictors)))

### Visualize Relationship
# Scatterplots for predictors in the test set
for (predictor in predictors) {
  p <- ggplot(eda_test_data, aes_string(x = predictor, y = response)) +
    geom_point(alpha = 0.7) +
    geom_smooth(method = "lm", color = "blue", se = FALSE) +
    labs(title = paste("Scatterplot of", response, "vs", predictor, "(Test Set)"),
         x = predictor, y = response) +
    theme_minimal()

  print(p)  # Explicitly print the plot
}

### Correlation Analysis
# Correlation matrix for the test set
cor_matrix_test <- cor(eda_test_data, use = "complete.obs")
print(cor_matrix_test)
corrplot(cor_matrix_test, method = "circle", type = "upper", tl.col = "black", tl.srt = 45)

### Linear Regression Model
# Fit the model using the test set
model_test <- lm(as.formula(paste(response, "~", paste(predictors, collapse = " + "))), data = eda_test_data)
summary(model_test)

# Extract residuals and fitted values from the test model
residuals_test <- resid(model_test)
fitted_values_test <- fitted(model_test)

# Calculate Variance Inflation Factors (VIF) for the test model
vif_test <- vif(model_test)
print(vif_test)

### Outlier Test Set
# Identify influential points using Cook's Distance
cooks_d_test <- cooks.distance(model_test)
influential_points_test <- which(cooks_d_test > 4 / (nrow(eda_test_data) - length(model_test$coefficients)))

# Output details of influential points
if (length(influential_points_test) > 0) {
  print(paste("Number of influential points:", length(influential_points_test)))
  print("Influential points identified:")
  print(influential_points_test)
} else {
  print("No influential points detected.")
}

```
Our analysis revealed that housing prices are influenced by various factors, such as absorption rates, unabsorbed homes, and economic indicators like GDP and CPI. Some predictors, like unabsorbed homes and GDP, have a strong and significant effect, while others, like construction starts, show weaker or inconsistent relationships. However, the strong connection between certain predictors, like GDP and CPI, makes it harder to separate their individual impacts, which complicates the analysis. By addressing these overlaps and focusing on the most reliable factors, we can build a clearer and more accurate model for understanding housing price trends.

## **3. Model Building**
- **Tools Used**: `lm` in R for linear regression modeling.  
- **Steps Taken**:  
  - Conducted all model building, assessments, and diagnostics on the **train_data** to ensure the model generalizes well to unseen data.  
  - Included key predictors based on theoretical relevance and statistical significance.  
  - Applied transformations to predictors to improve linearity and stabilize variance.  
  - Addressed residual skewness with a Box-Cox transformation of the response variable.  
  - Selected the final model based on diagnostics from the training set, ensuring adherence to regression assumptions and a good fit.

```{r}
# Load necessary libraries
library(car)      # For VIF calculation
library(MASS)     # For Box-Cox transformation
library(lmtest)   # For diagnostic tests

# Transform predictors
train_data$Detached_Absorption_Quarterly_Avg <- log(train_data$Detached_Absorption_Quarterly_Avg + 1)
train_data$Starting_Detached_Construction <- log(train_data$Starting_Detached_Construction + 1)
train_data$Under_Construction_Detached <- log(train_data$Under_Construction_Detached + 1)
train_data$Completed_Construction_Semi <- log(train_data$Completed_Construction_Semi + 1)
train_data$Detached_Unabsorbed_Quarterly_Avg <- log(train_data$Detached_Unabsorbed_Quarterly_Avg + 1)

# Box-Cox transformation
bc <- boxcox(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg +
                                Starting_Detached_Construction +
                                Detached_Unabsorbed_Quarterly_Avg +
                                Under_Construction_Detached +
                                Completed_Construction_Semi +
                                CPI_Quarterly_Avg,
             data = train_data, plotit = FALSE)
lambda <- bc$x[which.max(bc$y)]

if (lambda == 0) {
  train_data$Quarterly_Average <- log(train_data$Quarterly_Average)
} else {
  train_data$Quarterly_Average <- (train_data$Quarterly_Average^lambda - 1) / lambda
}

# Fit the linear regression model
model <- lm(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg +
                                   Starting_Detached_Construction +
                                   Detached_Unabsorbed_Quarterly_Avg +
                                   Under_Construction_Detached +
                                   Completed_Construction_Semi +
                                   CPI_Quarterly_Avg,
            data = train_data)

# multicollinearity
vif_values <- vif(model)
print("VIF Values:")
print(vif_values)

# final model
print("Final Model Summary:")
summary(model)

```

### Summary of Results:

The multiple linear regression model for predicting the Box-Cox transformed quarterly average housing prices (\(bc\_Quarterly\_Average\)) shows a strong overall fit (\(R^2 = 0.9839\), \(Adj. R^2 = 0.9806\), \(F\)-statistic \(p < 2.2e-16\)), indicating that the selected predictors explain nearly all variability in the response variable. Key significant predictors include **Detached_Unabsorbed_Quarterly_Avg** (\(p = 8.58e-05\), positive effect), **Under_Construction_Detached** (\(p = 9.90e-06\), positive effect), and **CPI_Quarterly_Avg** (\(p = 2.34e-11\), strong positive effect). **Detached_Absorption_Quarterly_Avg** (\(p = 0.0293\)) exhibits a significant negative relationship with prices. While **Starting_Detached_Construction** (\(p = 0.0831\)) and **Completed_Construction_Semi** (\(p = 0.1387\)) are less significant, their inclusion is supported by theoretical relevance.

### Justification for Transformations:

Transformations were applied to stabilize variance, improve linearity, and address multicollinearity. Log-transformations were used for predictors like **Detached_Absorption_Quarterly_Avg**, **Starting_Detached_Construction**, and **Under_Construction_Detached** to handle skewness and make relationships with the response variable more linear. Standardizing **CPI_Quarterly_Avg** addressed extreme multicollinearity, as it was highly correlated with other predictors. A Box-Cox transformation was applied to the response variable to correct residual skewness and ensure homoscedasticity. These steps collectively improved model diagnostics, reduced variance inflation factors (VIFs), and ensured compliance with regression assumptions, resulting in a more robust and interpretable model.

## **4. Assessment of Assumptions**
- **Tools Used**: Residual diagnostics using residual vs. fitted plots, Q-Q plots, and histograms.  
- **Findings**:  
  - **Linearity**: Residual plots indicated clustering around fitted values, with minor deviations suggesting potential non-linearity.  
  - **Homoscedasticity**: Fanning in residuals for predictors
  - **Normality**: Q-Q plots looks for alignment with normality
  - **Influential Points**

```{r}
# Load necessary libraries
library(ggplot2)
library(car)
library(MASS)
library(lmtest)

# Calculate residuals and fitted values
residuals <- resid(model)
fitted_values <- fitted(model)

# Residuals vs Fitted Values Plot 
ggplot(data = NULL, aes(x = fitted_values, y = residuals)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

# Residuals vs Each Predictor
predictors <- c("Detached_Absorption_Quarterly_Avg", 
                "Starting_Detached_Construction", 
                "Detached_Unabsorbed_Quarterly_Avg", 
                "Under_Construction_Detached", 
                "Completed_Construction_Semi", 
                "CPI_Quarterly_Avg")

for (predictor in predictors) {
  p <- ggplot(data = train_data, aes_string(x = predictor, y = "residuals")) +
    geom_point(alpha = 0.7) +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
    labs(title = paste("Residuals vs", predictor),
         x = predictor,
         y = "Residuals") +
    theme_minimal()

  print(p) 
  }

# Normal Q-Q Plot
qqnorm(residuals, main = "Normal Q-Q Plot")
qqline(residuals, col = "red")

par(mfrow = c(2, 2)) 
plot(model)            
par(mfrow = c(1, 1))

# Cook's Distance for Influential Points
cooksd <- cooks.distance(model)
plot(cooksd, type = "h", main = "Cook's Distance",
     xlab = "Observation", ylab = "Cook's Distance")
abline(h = 4/(nrow(train_data) - length(coef(model))), col = "red", lty = 2)

# VIF for Multicollinearity
vif_values <- vif(model)
print("Variance Inflation Factor (VIF) Values:")
print(vif_values)

```

The diagnostic analysis of the model shows that it performs well overall, but there are a few areas for improvement. Most predictors have acceptable relationships with the outcome variable, and the errors (residuals) are mostly normal, as shown in the Q-Q plot. However, some variables, like **CPI_Quarterly_Avg** and **log_Under_Construction_Detached**, show signs of overlap or dependency (multicollinearity), which could affect the stability of the results. A slight unevenness in the spread of errors (heteroscedasticity) suggests that adjustments, such as transformations, might improve the model's accuracy. One observation (Observation 10) has a bit more influence on the model than others, but it's not severe. Overall, the model is strong and reliable, but fine-tuning and closer inspection of specific predictors could make it even better.

## **5. Model Diagnostics**
- **Tools Used**: Performance metrics such as \( R^2 \), Adjusted \( R^2 \), AIC, BIC, and RMSE.  
- **Model Fit**:  
  - **\( R^2 \)**:
  - **Adjusted \( R^2 \)**
  - **AIC/BIC**:  
```{r}
# Load necessary libraries
library(Metrics)  # For RMSE calculation

# Model Performance Metrics
model_summary <- summary(model)

# Extract R^2 and Adjusted R^2
r_squared <- model_summary$r.squared
adj_r_squared <- model_summary$adj.r.squared

# AIC and BIC
aic <- AIC(model)
bic <- BIC(model)

# RMSE
residuals <- residuals(model)
rmse <- sqrt(mean(residuals^2))

# Results
cat("Model Performance Metrics:\n")
cat("R-squared: ", round(r_squared, 4), "\n")
cat("Adjusted R-squared: ", round(adj_r_squared, 4), "\n")
cat("AIC: ", round(aic, 4), "\n")
cat("BIC: ", round(bic, 4), "\n")
cat("RMSE: ", round(rmse, 4), "\n")

```

## **6. Model Validation**

```{r}

# Step 1: Use the model to predict on test_data
predictions <- predict(model, newdata = test_data)

# Step 2: Calculate residuals
residuals <- test_data$Quarterly_Average - predictions
# Replace `Quarterly_Average` with the actual response variable name in test_data

# Step 3: Compute Mean Squared Error (MSE) for test_data
mse_test <- mean(residuals^2)
cat("Mean Squared Error (MSE) on Test Data:", mse_test, "\n")

# Step 4: Compute R-squared for test_data
ss_total <- sum((test_data$Quarterly_Average - mean(test_data$Quarterly_Average))^2)
ss_residual <- sum(residuals^2)
r_squared_test <- 1 - (ss_residual / ss_total)
cat("R-squared on Test Data:", r_squared_test, "\n")

# Step 7: Print validation results
cat("Validation Results:\n")
cat("Mean Squared Error (Test Data):", mse_test, "\n")
cat("R-squared (Test Data):", r_squared_test, "\n")

```

## **7. Conclusion of Methods**
